---
title: "ShinyPractice0103"
format: html
execute: 
  warning: false
server: shiny
---

## Shiny Documents

This Quarto document is made interactive using Shiny. Interactive documents allow readers to modify parameters and see the results immediately. Learn more about Shiny interactive documents at <https://quarto.org/docs/interactive/shiny/>.

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change. This demonstrates how a standard R plot can be made interactive:

```{r setup}
library(shiny)
library(tidyverse)

dat <- read_csv("publish_practice.csv")





checkboxGroupInput(
  inputId = "checked_groups",
  label = "which groups do you want to display?",
  choices = c("a", "b", "c"),
  selected = c("a","b","c")
)

plotOutput("scatter")

```

```{r}
#| context: server
output$scatter <- renderPlot({
  plot_dat <- filter(dat, Group %in% input$checked_groups)
  
  ggplot(
    dat = plot_dat,
    aes(x = varX, y = varY, color = Group)
  ) + 
    geom_point()
})

```
