---
title: "PS1_First Term_Displaying and Thinking about public health data EDA (Exploratory Data Analysis)"
author: "Hyundam (Dami) Gu"
format: html
editor: visual
bibliography: references.bib
toc: true
toc-depth: 1
execute: 
  warning: false
code-annotations: hover
---

# **Problem 1. Air Pollution and Mortality in Baltimore**

## Background

Fossil fuel combustion releases particles and oxides that can form secondary particles, with historical evidence from the early 20th century showing that extreme air pollution can cause death. After Western countries reduced air pollution levels, public health researchers initially shifted their focus to morbidity studies, believing the mortality issue was resolved. However, recent studies, including research on daily variations in mortality, have shown significant links between particulate concentrations and death rates in some US regions. The NMMAPS study examined this relationship across 100 US cities, with specific data from Baltimore comparing daily deaths following the 20 highest and 20 lowest PM10 pollution days during summer months from 1987-2005.

[@peng2005]

## Data analysis

now, let's download this data and navigate this.

```{r}
library(tidyverse)
library(psych)
ps1 <- read_csv("baltps1.csv")
str(ps1)
```

let's create stem and leaf displays by group

```{r}
ps1.1 = filter(ps1, group==1)
ps1.2 = filter(ps1, group==2)
stem(ps1.1$deaths, scale=1)
stem(ps1.2$deaths, scale=1)
summary(ps1.1$deaths)
describe(ps1.1$deaths)
summary(ps1.2$deaths)
describe(ps1.2$deaths)
```

let's create a box and whiskers plot by group:

```{r}
boxplot(deaths ~ group, data=ps1)
```

## Data analysis

On high pollution days, the median daily death count 19 people, versus 17.5 people on low pollution days. The higher pollution days have wider range of daily deaths compare to low pollution days, 23 deaths for higher pollution days and 16 deaths for low pollution days, suggesting more variability in mortality. Both distributions are slightly skewed to the right. These findings are consistent with the hypothesis that increased particulate pollution is associated with higher daily mortality rates in the studied area.

# Problem 2. Costs of Carotid Endarterectomy in Maryland

## Background

Carotid endarterectomy (CE) is a vascular surgical procedure performed to improve blood flow through the carotid artery and reduce the risk of stroke and sudden death, with approximately 2,000 procedures performed annually across Maryland's 50+ hospitals. The State of Maryland Health Services Cost Review Commission (HSCRC) collects and makes public the data for each procedure, which can be used to analyze whether increased surgical experience (both by physician and hospital) correlates with reduced risk of post-surgical stroke or death. This data from 1990-1995 will be used to explore the distribution of procedure costs across Maryland's CE procedures.

[@warlow1991a]

## Data Processing

Let's download the data and navigate it, with making stem and leaf plots of the male and female CE costs.

```{r}
library (tidyverse)
ce621<-read_csv("ce621.csv")
str(ce621)
ce621.male = filter(ce621, sex=="Male")
ce621.female = filter(ce621, sex=="Female")
stem(ce621.male$totchg) #<1>
stem(ce621.female$totchg) 
```

1.  According to codebook, totchg is Total charges (dollars).

Let's make the summary table regarding CE costs by sex.

```{r}
summary(ce621.male$totchg)
sd(ce621.male$totchg)
quantile(ce621.male$totchg, c(0,.1,.25,.5,.75,.9,1))
summary(ce621.female$totchg)
sd(ce621.female$totchg)
quantile(ce621.female$totchg, c(0,.1,.25,.5,.75,.9,1))
```

| Statistics      | Male                 | Female                 |
|-----------------|----------------------|------------------------|
| *Typical Value* |                      |                        |
| Mean (X)        | \$ 6484              | \$8099                 |
| Median          | \$ 5597.5            | \$5643.5               |
| *Variability*   |                      |                        |
| IQR             | 7189-4327.5=\$2861.5 | 9546.25-4359.25=\$5187 |
| SD              | \$3278               | \$6679                 |

Now let's look at the specific year of 1995, and make a sex comparison within age group of ≤ 50 years; 51-64; 65+. We will do this using a logarithmic (base 10) scale for dollars as well.

```{r}
ce621e <- read_csv("ce621entire.csv")
ce621e95 = filter(ce621e, year==1995)
ce621e95$agecat = cut(ce621e95$age, c(0, 50, 64, 100), right=TRUE, labels=c("<=50","51-64",">=65"))
boxplot(totchg ~ sex + agecat, data=ce621e95, names=c("F <= 50","M <= 50","F 51-64", "M 51-64", "F >=65", "M >=65"))
```

```{r}
ce621e95$log10chg = log10(ce621e95$totchg)
boxplot(log10chg ~ sex + agecat, data=ce621e95,
names=c("F <= 50", "M <= 50","F 51-64", "M 51-64","F >=65", "M >=65"))
```

## Data analysis

A sample of 200 carotid endarterectomy operations in Maryland, including surgeries performed on 100 men and 100 women, permit evaluation of the health care burden of this type of operation.

-   The mean cost of such a surgery was much higher for women than for men (\$8099 vs.\$6484, respectively).

-   The median values were comparable (about \$5600 for both sexes) suggesting that some women’s surgeries are far more expensive than the majority of the sample.'

-   The maximum value for women is about \$40083, as compared to less than half that much for men.

-   The charges for women are more variable throughout the range. SD is about twice that seen among the men, and the IQR is wider among females.

These data suggest that while the lower end of the price spectrum is similar for both sexes, there was a group of women who paid a lot more money than did men in our sample. Thus the data for women are more skewed towards higher values than those for men.

In the data of year 1995, the cost of carotid endarterectomy varies by age and sex in this sample. The median value is about the same across groups, except that the data for women under age 51 showed a higher median and are more variable as reflected by the wider interquartile range. The data for both men and women aged above 50 are skewed towards high values; though women have higher extreme values than men within each age stratum. The spread of the data increases with age.

When graphed on a log 10 scale, the distribution of charges all appear less skewed. Variability in surgical charge increases with age in both sexes, though the change is more dramatic among men. The log 10 transformation has the effect of decreasing the value of extremely high observations more than it changes lower values, so that outlying high values are pulled in towards the rest of the data.
